---
title: "Creating an R Package"
subtitle: "An introduction to creating your own R package"
author: '[Dominique Maucieri](https://www.dominiquemaucieri.com)'
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
  pdf_document:
    toc: yes
    toc_depth: 3
---

# Setup

Welcome to this tutorial on how to create an R package. To begin you will need to ensure you have all the needed packages installed, and then load the packages. 

```{r install packages, eval = FALSE}

install.packages(c("devtools", "roxygen2", "usethis", "testthat", "covr"))

```

```{r load packages, eval = FALSE}

library(devtools)
library(roxygen2)
library(usethis)
library(testthat)
library(covr)

```
These are the packages needed to build an R package, though if your package functions will depend on other functions, you will need to add those as you go as well.

# The Name

Determining the name of your package can be very challenging, as it will be how your package will be known to everyone that uses your functions.

Time and care should be put into your package name choice. The available package has a great function called available() which you can use to ensure that there isn't already a package with the name you want or unintentional meanings to your package name. 

Your name must fit 3 criteria:

  1. Only letters, numbers and periods (though periods are not recommended)
  2. Start with a letter
  3. Cannot end with a period

We are going to create an example package called standarderror, so you can practice checking that package name with this code:

```{r available, eval=FALSE}

install.packages("available")
library(available)

available::available("standarderror")

```

This shows that "standarderror" is valid on all platforms and nothing was found on any of the online databases either. So this is a good name to continue with.

# Creating the package skeleton

To create the skeleton of your new package, you only need one line of code. You will need to change `/path/to` to the location path were you want this package on your computer

```{r package skeleton, eval=FALSE}

usethis::create_package("~/path/to/standarderror")

```

Looking at this directory, there are a few files that were created

- *.Rbuildignore*

  * Files we need but will not be included when building the package

- *.gitignore*

  * Ignores some files created by R and RStudio
  * Harmless
  
- *R/*

  * Folder containing the .R files
  
- *standarderror.Rproj*

  * RStudio project
  
- *.Rproj.user*

  * May not have
  
  * A directory used by RStudio internally
  
- *DESCRIPTION*

  * Package metadata
  
- *NAMESPACE*

  * Declares your package exports and the imports from other packages
  
  * DO NOT EDIT
  
  
## Editing the DESCRIPTION file

At first your description file will look something like this:

Package: standarderror \newline
Title: What the Package Does (One Line, Title Case) \newline
Version: 0.0.0.9000 \newline
Authors@R:\newline
             *First Last <first.last@example.com> [aut, cre] (YOUR-ORCID-ID) \newline
Description: What the package does (one paragraph). \newline
License: `use_mit_license()`, `use_gpl3_license()` or friends to pick a
    license \newline
Encoding: UTF-8 \newline
Roxygen: list(markdown = TRUE) \newline
RoxygenNote: 7.1.2 \newline


But lets update this so it is more functional. The package name looks great but lets add this title: Calculation of Standard Error. The version number ends in 9000 which signifies this package is in development. Next you can edit the Author section with your name email and ORCID if you have one, but I will include how I have filled it out so you know what to add. If you were creating a package with collaborators you would also add them to this spot too, there are a lot of resources online for adding more authors and people with other roles. We can also add a description of what this package will do and add a license. For this example we will use the MIT license but for your own work you should look into which license is best for you.

```{r license, eval = FALSE}

usethis::use_mit_license()

```

This should leave us with a DESCRIPTION file that looks something like this (but with your name):

Package: standarderror \newline
Title: Calculation of Standard Error \newline
Version: 0.0.0.9000 \newline
Authors@R:  \newline
person("Dominique", "Maucieri", role = c("aut", "cre"), email = "dominiquemaucieri@gmail.com",
           comment = c(ORCID = "0000-0003-1849-2472")) \newline
Description: This package will aid in calculation of the standard error of means. \newline
License: MIT + file LICENSE \newline
Encoding: UTF-8 \newline
Roxygen: list(markdown = TRUE) \newline
RoxygenNote: 7.1.2


# Creating a function

We are going to create the function std.err(). To create a new function, you use the use_r() function from the usethis package in your console. It will create the file for you and open it.

```{r create function, eval = FALSE}

usethis::use_r("std.err")

```

Now in this blank .R file, we will add the code for our function. If you are unfamiliar with how to code functions, there are many great resources online including [Functions in Advanced R](https://adv-r.hadley.nz/functions.html) and [this tutorial from our Coding Club](https://ourcodingclub.github.io/tutorials/funandloops/)

We want our function to receive a vector of numbers and calculate the standard error of the mean of that vector. To calculate standard error, it will be the standard deviations divided by the square-root of the sample size, which written as a function will look like this:

```{r fill function, eval = FALSE}

std_err <- function(x){

  x.std.err <- sd(x) / sqrt(length(x))

  return(x.std.err)

}

```


## Function Documentation

Now this is only the function, there is no associated information about the elements of this function, like what x is. To add this extra information in, you need to have your cursor within the function you are working on, then select Code>Insert Roxygen Skeleton. Now you just have to fill it in. For std_err() we will fill in the documentation like:

```{r roxygen skeleton, eval = FALSE}

#' @title Calculate standard error
#' @description Using a vector of samples, calculate the standard error of the
#'     mean.
#'
#' @param x A vector of numerical data.
#'
#' @return A numeric value.
#' @export
#'
#' @examples
#' std_err(x = c(4, 7, 2, 7, 5, 9))
#'
std_err <- function(x){

  x.std.err <- sd(x) / sqrt(length(x))

  return(x.std.err)

}

```

When filling out this documentation, you should keep each row around 80 characters long. After that you will have to manual hit return and continue writing on a new line. When you start a new line, there needs to be 4 spaces at the start which will make sure when we build this documentation manual, it knows that this line is a continuation of the previous one. You can see this in the description. How to know when 80 characters is? I it is not already there, you can add a margin line at 80 characters by going to Preferences > Code > Display > Show Margin. 

Additionally when adding examples, you can add as many as you want, ensuring the code will work and run without warnings. If you want to include code with errors, you can wrap it with `\dontrun{     }`

## Package Dependancies

Sometimes you may want to use a functions that already exist from other packages within your functions. Be cautious on how many and which ones you use as your code will rely on other peoples code. It is a good idea to get into the habit of specifying the package that every function you use in your code is from. So for `std_err()` both the `sqrt()` and `length()` functions are from base R, however, `sd()` is from the stats package. While the stats package is considered a base package as well, it is a good idea to still specify that you are using `stats::sd()` within your function. 

```{r roxygen skeleton with package, eval = FALSE}

#' @title Calculate standard error
#' @description Using a vector of samples, calculate the standard error of the
#'     mean.
#'
#' @param x A vector of numerical data.
#'
#' @return A numeric value.
#' @export
#'
#' @examples
#' std_err(x = c(4, 7, 2, 7, 5, 9))
#'
std_err <- function(x){

  x.std.err <- stats::sd(x) / sqrt(length(x))

  return(x.std.err)

}

```

For other packages that are not base packages, you will have to add a dependency. There are a few different types, the most common are "Suggests" and "Imports". 

- Suggests will be for packages or functions that are optional

  * only used in a single function
  * only used in vignettes
  
- Imports will be for packages or functions that are not optional for the working of you package

For example if you wanted to use the dplyr package with your functions you would specify in your command line:

```{r dependency, eval = FALSE}

usethis::use_package("dplyr", type = "Imports")

```

And you would only have to do this one and it will be added to your DESCRIPTION file and available for use in your functions. After that any time you use a dplyr function just put `dplyr::` before the function name.

One exception to this is if you want to use the pipe function `%>%`. This function has special documentation and you don't specify the package before you use the %>%, you just use it as is in your code. To be able to use it you must specify first and only once, into your console: 

```{r pipe, eval = FALSE}

usethis::use_pipe()

```

Now that adding packages and documentation to our function is done, you can run the following code in your console, to write the manual file and add your function to be exported to the NAMESPACE file.

```{r document, eval = FALSE}

devtools::document()

```

Now you will have a new folder in your package directory called man/ and within that there is an R Documentation file called std_err.Rd. This file is not editable by hand, to change it you have to change your documentation in the roxygen skeleton. If you select the preview button at the top of the .Rd file, you can view the manual for your function. This is the same information that would be available if you searched for the help on this function or typed `?std_err` into the console. 


# Does it all work?

Now that we have a function that is properly documented, we are probably wondering if the changes we have made work. There are a few different things you can do to examine your package. 

```{r load_all, eval=FALSE}

load_all()

std_err(x = c(4, 7, 2, 7, 5, 9))

```

`load_all()` simulates building and installing your package. Once loaded you can use work through an example to check that package works.

```{r check, eval=FALSE}

check()

```

You should use `check()` often to ensure that there are no errors within your package. It is easier to fix issues when you know the exact code you changed that produced them, instead of having to work through an entire package to find the issue. At the end of the `check()` output, there will be a summary of errors, warnings and notes, the goal is to have 0. Sometimes there may be notes that you can not get rid of but try your best to eliminate them. We will discuss the testthat package in a moment, but when you use `check()` it will also run through all your tests to ensure your package work the way you intend it too as well.

```{r install, eval=FALSE}

install()

```

Finally, `install()` will build and install your package into your library, so that you can use it as you would with any package.


# Function Tests

Each function should have a test made for it, to ensure that while there are no errors with the code, that the code is actually doing what you think it is doing. For our example package we will create a tests for `std_err()` 

```{r test creation, eval = FALSE}

usethis::use_test("std_err")

```

Executing this code in the console will create a few things.

1. Create new folders test/testthat/

    * this folder will hold all your tests
  
2. Creates tests/testthat.R
  
    * this document will be run with each use of `check()` to check that all the tests work properly
  
    * you don't need to change this by hand as long as you create every test with `usethis::use_test()`
  
3. Creates tests/testthat/test-std_err.R

    * this file we will edit to create the specific tests for the `std_err()`
  
 
Every new test .R file will start with the same code:

```{r example with multiplication, eval = FALSE}

test_that("multiplication works", {
  expect_equal(2 * 2, 4)
})

```

This is just an example test, as you can see there is nothing here that will test our function. We will change this so that we add in an example and test that the example will yield the result that we are expecting. My preference is to use the same example that I used in the package documentation, but you can do more or different examples, its up to you. You just want to make sure that you test all aspects or options within your function thoroughly. 

Here is the test for `std_err()` work through the code to see if it makes sense to you and you see how we are testing it.

```{r actual test, eval = FALSE}

example_data <- c(4, 7, 2, 7, 5, 9)

example_se <- std_err(x = example_data)

expect_equal(example_se, 1.02198064778)

```

After creating and saving this test, you can test it with the following code:

```{r conducting test, eval=FALSE}

devtools::test()

```

This will tell you the status of all your function tests so you can make sure that all your functions are working just as you want them to.

## Coverage

How much of your code is being used in tests. This should be 100% currently as we only made one function and we wrote a test for it. As you continue to make functions, you want this to remain high so you will have to keep making more functions.Low coverage could mean you miss issues.

```{r coverage, eval = FALSE}

covr::package_coverage()

```


# Other aspects

## Vignettes

```{r vignette, eval=FALSE}

usethis::use_vignette("vignette-name")

```

Using this code will create a vignette document, which is a form of long form documentation, kind of like a tutorial. You can use it to show the full functionality of your package or work through different examples. This code produced an .Rmd file which you can just edit to produce the vignette you want. 


## GitHub

If you are unfamiliar with GitHub, I recommend you look into it as it is a great way to share code and even packages. For example, as this package is on my GitHub, you can download this package (which is the same as using install.package() but for a package on GitHub) with:

```{r GitHub, eval = FALSE}

devtools::install_github("DominiqueMaucieri/standarderror")

```

Using GitHub is also a benefit as it can allow you to check other platforms with GitHub actions allowing for continuous integration and you can set GitHub to check your coverage. 

## Package Website

Once your package is on GitHub, a website can be nice to help people get oriented and familiar with your package. It is easy to set this up with pkgdown.

```{r pkgdown, eval=FALSE}

usethis::use_pkgdown_github_pages()

```

## Adding data

Adding data can be helpful for your package as well, and if you want to do this it is something you should look into more. There are typically 3 steps to adding data to your package:

1. Add raw data to a data-raw/ folder

```{r raw data, eval=FALSE}

usethis::use_data_raw("dataname")

```

2. Then you will either use that script to import data or create the data. After which you will save the raw data as data:

```{r data, eval=FALSE}

usethis::use_data(dataname, overwrite = TRUE)

```

3. Then you will need to add documentation for the data set. This differs from function documentation so take some time to familiarize yourself with how to do this. 


Then once your package is built and installed, you will be able to load the data with:

```{r load data, eval=FALSE}

data("dataname")

```


And that is the basics and some future directions. There is a lot more information out there on creating packages so I hope this was able to inspire you and show you that you have the skills to be able to create a package.
